Fast gravitational wave parameter estimation without compromises

### 2023-02-08 src/tex/ms.tex
Update

% Abstract with filler text
\begin{abstract}
\te{There is still a "KW:fill" remaining in the text!}
We present a lightweight, flexible, and high-performance framework for inferring
the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible waveforms,
and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced by
normalizing flows, we achieve full Bayesian parameter estimation for real events
like GW150914 and GW170817 within a minute of sampling time.  Our framework does
not require pre-training or explicit reparameterizations and can be generalized
to handle higher dimensional problems. We present the details of our
implementation and discuss trade-offs and future developments in the context of
other proposed strategies for real-time parameter estimation. Our code for
generating the manuscript and running the analysis is publicly available on
GitHub. \te{Add link to Github}
\end{abstract}

@@

To compute the integral in Eq.~\eqref{eq: innerproduct}, we need to evaluate a
waveform model $h(\mathbf{\theta})$ at a number of frequency bins. This makes
evaluating the likelihood function often the most computationally intensive
part of PE. The most accurate waveforms are obtained via numerical relativity
(NR) simulations \citep{BaumgarteShaprio:NumRel}, which directly solve the Einstein equations numerically for a
given system. Unfortunately, such simulations can take anywhere from a day to
half a year, making direct evaluation through NR prohibitively expensive. To
circumvent this, there are several families of waveform ``approximants'',
including the \textsc{IMRPhenom} family \citep{Khan:2015jqa,
Garcia-Quiros:2020qpx}, the \textsc{SEOB} family \citep{PhysRevD.89.061502}, and
the NR surrogate family \citep{Varma:2019csw}. Evaluating those approximants can
take ${\sim}10$ ms to 1 s for short signals, such as a $30-30\ M_{\odot}$
binary black hole, reaching up to \kw{Fill} for long signals, such as a
$1.4-1.4\ M_{\odot}$ binary neutron star.  Since PE requires millions of
likelihood evaluations during sampling, the computational cost in evaluating
the waveform is a major contributor to the long runtimes of GW PE.

@@

To further quantify how well our result agrees with a uniform distribution, we
can compute the Kolmogorov-Smirnov $p$-values for each marginalized
distribution \citep{}\te{There is apparently supposed to be a citation here?}.
A low $p$-value (with a threshold often
chosen to be $p = 0.05$) could indicate that our result is in tension with a uniform
distribution. The $p$-values obtained for each parameter are shown in the
legend of Fig.~\ref{fig:ppplot}. Most of them are well above the $p = 0.05$
threshold, except for $\psi$, which is close to the threshold. Once
again, assuming these $p$-values are drawn from a uniform distribution, given
11 draws (the number of parameters in our inference), it is not abnormal to
have one of the parameters lying slightly outside the threshold. To assess
whether this is expected, we can compute the combined $p$-value for these 11
parameters, and find it to be $p = \ptot$.  This shows our inference pipeline
performs properly on simulated data at a similar level as standard tools
\citep{Veitch:2014wba,Romero-Shaw:2020owr}.

@@

\begin{figure*}
    \includegraphics[width=0.99\linewidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange). \te{Change font to be latex, make labels larger, make figure full page.}
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\linewidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange). \te{Same as Fig 2}
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files included in our data release \citep{release} \te{Add citation}.  We
carry out these runs using \textsc{parallel Bilby} (\textsc{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

>>>>

% Abstract with filler text
\begin{abstract}
We present a lightweight, flexible, and high-performance framework for inferring
the properties of gravitational-wave events. By combining likelihood
heterodyning, automatically-differentiable and accelerator-compatible waveforms,
and gradient-based Markov chain Monte Carlo (MCMC) sampling enhanced by
normalizing flows, we achieve full Bayesian parameter estimation for real events
like GW150914 and GW170817 within a minute of sampling time.  Our framework does
not require pre-training or explicit reparameterizations and can be generalized
to handle higher dimensional problems. We present the details of our
implementation and discuss trade-offs and future developments in the context of
other proposed strategies for real-time parameter estimation. Our code for running the analysis is publicly available on
GitHub \url{https://github.com/kazewong/jim}.
\end{abstract}

@@

To compute the integral in Eq.~\eqref{eq: innerproduct}, we need to evaluate a
waveform model $h(\mathbf{\theta})$ at a number of frequency bins. This makes
evaluating the likelihood function often the most computationally intensive
part of PE. The most accurate waveforms are obtained via numerical relativity
(NR) simulations \citep{BaumgarteShaprio:NumRel}, which directly solve the Einstein equations numerically for a
given system. Unfortunately, such simulations can take anywhere from a day to
half a year, making direct evaluation through NR prohibitively expensive. To
circumvent this, there are several families of waveform ``approximants'',
including the \textsc{IMRPhenom} family \citep{Khan:2015jqa,
Garcia-Quiros:2020qpx}, the \textsc{SEOB} family \citep{PhysRevD.89.061502}, and
the NR surrogate family \citep{Varma:2019csw}. Since PE requires millions of
likelihood evaluations during sampling, the computational cost in evaluating
the waveform is a major contributor to the long runtimes of GW PE.

@@

To further quantify how well our result agrees with a uniform distribution, we
can compute the Kolmogorov-Smirnov $p$-values for each marginalized
distribution \citep{doi:10.1080/01621459.1968.11009335}.
A low $p$-value (with a threshold often
chosen to be $p = 0.05$) could indicate that our result is in tension with a uniform
distribution. The $p$-values obtained for each parameter are shown in the
legend of Fig.~\ref{fig:ppplot}. Most of them are well above the $p = 0.05$
threshold, except for $\psi$, which is close to the threshold. Once
again, assuming these $p$-values are drawn from a uniform distribution, given
11 draws (the number of parameters in our inference), it is not abnormal to
have one of the parameters lying slightly outside the threshold. To assess
whether this is expected, we can compute the combined $p$-value for these 11
parameters, and find it to be $p = \ptot$.  This shows our inference pipeline
performs properly on simulated data at a similar level as standard tools
\citep{Veitch:2014wba,Romero-Shaw:2020owr}.

@@

\begin{figure*}
    \includegraphics[width=0.99\textwidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (grey contour) and \textsc{Bilby} (blue line).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\textwidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (grey contour) and \textsc{Bilby} (blue line).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github \url{https://github.com/kazewong/TurboPE/tree/main/src}.  We
carry out these runs using \textsc{parallel Bilby} (\textsc{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.
